Staged Compilation With Two-Level Type Theory

### 2022-06-17 paper/paper.tex
revision, add appendix

\documentclass[acmsmall,anonymous,review]{acmart}
%% \documentclass[nonacm,acmsmall,review]{acmart}
%% \raggedbottom

@@

\begin{abstract}
  The aim of staged compilation is to enable metaprogramming in a way such that
  we have guarantees about the well-formedness of code output, and we can also
  mix together object-level and meta-level code in a concise and convenient
  manner. In this work, we observe that two-level type theory (2LTT), a system
  originally devised for the purpose of developing synthetic homotopy theory,
  also serves as a system for staged compilation with dependent types. 2LTT has
  numerous good properties for this use case: it has a concise specification,
  well-behaved model theory, and it supports a wide range of language features
  both at the object and the meta level. First, we give an overview of 2LTT's
  features and applications in staging. Then, we present a staging algorithm and
  prove its soundness and stability. Our algorithm is ``staging-by-evaluation'',
  analogously to the technique of normalization-by-evaluation, in that staging
  is given by the evaluation of 2LTT syntax in a semantic domain. A sound and
  stable staging algorithm constitutes a proof of strong conservativity of 2LLT
  over the object theory. To our knowledge, this is the first description of
  staged compilation which supports full dependent types and unrestricted
  staging for types.
\end{abstract}

@@

\subsection{Properties of Lifting \& Binding Time Improvements}

@@

Fortunately, our specified meta type theory can be interpreted in
\emph{setoids}, as most recently explained by Pujet and Tabareau
\cite{DBLP:journals/pacmpl/PujetT22}. A setoid is a set together with an
equivalence relation on it. We interpret closed types as setoids, and
dependent types as setoid fibrations. For inductive base types such $\Bool$ or
$\Nat$, the equivalence relation is interpreted as identity. Functions are
interpreted as equivalence-preserving functions, and two functions are equivalent if
they are pointwise equivalent. A quotient type $A/R$ is interpreted by extending
the semantic notion of equivalence for $A$ with the given $R$ relation.

@@

In other words, we can view the usage of constructive quotients as a convenient
shorthand for working with explicit equivalence relations (an \emph{internal
language} of setoids). When we reason about the syntax of a type theory, this
provides a bridge between the categorical/algebraic style and the traditional
style that uses separate equivalence relations. It makes sense to work with
quotients internally to a type theory whenever we only want to talk about
constructions and properties which respect conversion. However, at any point we
have the option to switch to an external perspective and look at underlying sets
and the underlying maps.

@@

@@

The following notions are derivable:
\begin{itemize}
\item
  By moving left-to-right along $\Sub\,\Gamma\,(\Delta\ext A) \simeq ((\sigma :
  \Sub\,\Gamma\,\Delta) \times \Tm_{i,j}\,\Gamma\,(A[\sigma]))$,
  and starting from the identity morphism $\id : \Sub\,(\Gamma\ext A)\,(\Gamma\ext A)$, we recover
  the \emph{weakening substitution} $\p : \Sub\,(\Gamma\ext A)\,\Gamma$ and the \emph{zero variable}
  $\q : \Tm_{i,j}\,(\Gamma\ext A)\,(A[\p])$.
\item
  By weakening $\q$, we recover a notion of variables as De Bruijn
  indices. In general, the $n$-th De Bruijn index is defined as $\q[\p^{n}]$,
  where $\p^{n}$ denotes $n$-fold composition.
\item
  By moving right-to-left along $\Sub\,\Gamma\,(\Delta\ext A) \simeq ((\sigma :
  \Sub\,\Gamma\,\Delta) \times \Tm_{i,j}\,\Gamma\,(A[\sigma]))$, we recover the
  operation which extends a morphism with a term. In the initial model, this
  extends a parallel substitution with an extra term, thus justifying the view
  of substitutions as lists of terms. We denote the extension operation as
  $(\sigma,\,t)$ for $\sigma : \Sub\,\Gamma\,\Delta$.
\end{itemize}

\begin{notation}
De Bruijn indices are rather hard to read, so we will sometimes use nameful notation
for binders and substitutions. For example, we may write $\Gamma \ext (x : A)
\ext (y : B)$ for a context, and subsequently write $B[y \mapsto t]$ for
substituting the $x$ variable for some term $t : \Tm_{i,j}\,\Gamma\,A$. Using
nameless notation, we would instead have $B : \Ty_{i,j}\,(\Gamma \ext A)$ and
$B[\id,\,t]$; here we recover single substitution by extending the identity
substitution $\id : \Sub\,\Gamma\,\Gamma$ with $t$.

We may also use implicit weakening: if a type or term is in a $\Gamma$ context,
we may use it in an extended $\Gamma \ext A$ context without marking the
weakening substitution.
\end{notation}

\begin{definition}
A \textbf{model of 2LTT} is a model of basic 2LTT which supports certain type
formers. For the sake of brevity, we specify here only a small collection of
type formers. However, we will argue later that our results extend to more
general notions of inductive types. We specify type formers in the following. We
omit substitution rules for type and term formers; all of these can be specified in
an evident structural way, e.g.\ as in $(\Lift A)[\sigma] = \Lift(A[\sigma])$.
\begin{itemize}
\item \emph{Universes}. For each $i$ and $j$, we have a Coquand-style universe
  \cite{coquand2018canonicity} in $\Ty_{i,j}$. This consists of $\U_{i,j} : \Ty_{i,j+1}\,\Gamma$,
  together with $\El : \Tm_{i,j+1}\,\Gamma\,\U_{i,j} \to \Ty_{i,j}\,\Gamma$ and $\Code$, where $\Code$
  and $\El$ are inverses.
\item \emph{$\Sigma$-types}. We have $\Sigma\,(x : A)\,B : \Ty_{i,j}\,\Gamma$ for $A : \Ty_{i,j}\,\Gamma$
  and $B : \Ty_{i,j}\,(\Gamma \ext (x : A))$, together with a natural isomorphism of pairing and projections:
  \[ \Tm_{i,j}\,\Gamma\,(\Sigma\,(x : A)\,B) \simeq ((t : \Tm_{i,j}\,\Gamma\,A) \times \Tm_{i,j}\,\Gamma\,(B[x\mapsto t]) \]
  We write $(t,\,u)$ for pairing and $\fst$ and $\snd$ for projections.
\item \emph{Function types}.
  We have $\Pi\,(x : A)\,B : \Ty_{i,j}\,\Gamma$ for $A : \Ty_{i,j}\,\Gamma$ and
  $B : \Ty_{i,j}\,(\Gamma \ext (x : A))$, together with abstraction and application
  \begin{alignat*}{3}
    & \lam &&: \Tm_{i,j}\,(\Gamma \ext (x : A))\,B \to \Tm_{i,j}\,\Gamma\,(\Pi\,(x : A)\,B) \\
    & \app &&: \Tm_{i,j}\,\Gamma\,(\Pi\,(x : A)\,B) \to \Tm_{i,j}\,(\Gamma \ext (x : A))\,B
  \end{alignat*}
  such that $\lam$ and $\app$ are inverses. This specification of $\app$ is
  equivalent to the ``traditional'' specification:
  \[ \app' : (t : \Tm_{i,j}\,\Gamma\,(\Pi\,(x :
  A)\,B))\to (u : \Tm_{i,j}\,\Gamma\,A) \to \Tm_{i,j}\,\Gamma\,(B[x \mapsto
    u]).\]
  The traditional version is definable as $(\app\,t)[\id,\,u]$. We use
  the $\app$-$\lam$ isomorphism because it is formally more convenient.

\item \emph{Natural numbers}. We have $\Nat_{i,j} : \Ty_{i,j}\,\Gamma$, $\zero_{i,j} : \Tm_{i,j}\,\Gamma\,\Nat_{i,j}$,
  and $\suc_{i,j} : \Tm_{i,j}\,\Gamma\,\Nat_{i,j} \to \Tm_{i,j}\,\Gamma\,\Nat_{i,j}$. The eliminator is the following.
  \begin{alignat*}{5}
    & \NatElim :\hspace{0.5em}
                &&  (P &&: \Ty_{i,k}\,(\Gamma \ext (n : \Nat_{i,j}) \\
    &           &&  (z &&: \Tm_{i,k}\,\Gamma\,(P[n\mapsto \zero_{i,j}])) \\
    &           &&  (s &&: \Tm_{i,k}\,(\Gamma \ext (n : \Nat_{i,j}) \ext (\mit{pn} : P[n\mapsto n]))\,(P[n\mapsto \suc_{i,j}\,n]))\\
    &           &&  (t &&: \Tm_{i,j}\,\Gamma\,\Nat_{i,j}) \\
    &\hspace{3em}\to        &&&& \Tm_{i,j}\,\Gamma\,(P[n\mapsto t]))
  \end{alignat*}
  We also have the $\beta$-rules:
  \begin{alignat*}{5}
    & \NatElim\,P\,z\,s\,\zero_{i,j}     &&= z \\
    & \NatElim\,P\,z\,s\,(\suc_{i,j}\,t) &&= s[n \mapsto t,\,\mit{pn} \mapsto \NatElim\,P\,z\,s\,t]
  \end{alignat*}
  Note that we can eliminate from a level $j$ to any level $k$.
\end{itemize}
\end{definition}

@@

\subsubsection{The setoid interpretation of the syntax} We need to show
that the 2LTT syntax, as a particular inductive type, can be modeled in setoids.
Unfortunately, there is no prior work which presents the setoid interpretation
of quotient types in the generality of quotient induction-induction. We sketch
it for our specific use case, but we argue that the interpretation is fairly
mechanical and could be adapted to any quotient inductive-inductive type.
We focus on contexts and types here.

>>>>

\documentclass[acmsmall,review]{acmart}
%% \documentclass[nonacm,acmsmall,review]{acmart}
%% \raggedbottom

@@

\begin{abstract}
The aim of staged compilation is to enable metaprogramming in a way such that we
have guarantees about the well-formedness of code output, and we can also mix
together object-level and meta-level code in a concise and convenient manner. In
this work, we observe that two-level type theory (2LTT), a system originally
devised for the purpose of developing synthetic homotopy theory, also serves as
a system for staged compilation with dependent types. 2LTT has numerous good
properties for this use case: it has a concise specification, well-behaved model
theory, and it supports a wide range of language features both at the object and
the meta level. First, we give an overview of 2LTT's features and applications
in staging. Then, we present a staging algorithm and prove its correctness. Our
algorithm is ``staging-by-evaluation'', analogously to the technique of
normalization-by-evaluation, in that staging is given by the evaluation of 2LTT
syntax in a semantic domain. The staging algorithm together with its correctness
constitutes a proof of strong conservativity of 2LLT over the object theory. To our
knowledge, this is the first description of staged compilation which supports
full dependent types and unrestricted staging for types.
\end{abstract}

@@

\subsection{Properties of Lifting, Binding Time Improvements}

@@

Fortunately, our specified meta type theory can be interpreted in
\emph{setoids}, as most recently explained by Pujet and Tabareau
\cite{DBLP:journals/pacmpl/PujetT22}. A setoid is a set together with an
equivalence relation on it. We interpret closed types as setoids, and
dependent types as setoid fibrations. For inductive base types such $\Bool$ or
$\Nat$, the equivalence relation is interpreted as identity. Functions are
interpreted as equivalence-preserving functions, and two functions are equivalent if
they are pointwise equivalent. A quotient type $A/R$ is interpreted by extending
the semantic notion of equivalence for $A$ with the given $R$ relation. We will
sketch the interpretation of the 2LTT syntax (given as a quotient inductive-inductive type)
in Section \ref{sec:2lttsetoid}.

@@

In other words, we can view the usage of constructive quotients as a convenient
shorthand for working with explicit equivalence relations. When we reason about
the syntax of a type theory, this provides a bridge between the
categorical/algebraic style and the traditional style that uses separate
equivalence relations. It makes sense to work with quotients internally to a
type theory whenever we only want to talk about constructions and properties
which respect conversion. However, at any point we have the option to switch to
an external perspective and look at underlying sets and the underlying maps.

@@

We provide an appendix to this paper, where we present a listing of all rules
and equations of 2LTT in two styles: first, in a more traditional style with
derivation rules, and also in a very compact manner using a second-order
algebraic signature.

@@

The following notions are derivable:
\begin{itemize}
\item
  By moving left-to-right along $\Sub\,\Gamma\,(\Delta\ext A) \simeq (\sigma :
  \Sub\,\Gamma\,\Delta) \times \Tm_{i,j}\,\Gamma\,(A[\sigma])$,
  and starting from the identity morphism $\id : \Sub\,(\Gamma\ext A)\,(\Gamma\ext A)$, we recover
  the \emph{weakening substitution} $\p : \Sub\,(\Gamma\ext A)\,\Gamma$ and the \emph{zero variable}
  $\q : \Tm_{i,j}\,(\Gamma\ext A)\,(A[\p])$.
\item
  By weakening $\q$, we recover a notion of variables as De Bruijn
  indices. In general, the $n$-th De Bruijn index is defined as $\q[\p^{n}]$,
  where $\p^{n}$ denotes $n$-fold composition of $p$.
\item
  By moving right-to-left along $\Sub\,\Gamma\,(\Delta\ext A) \simeq (\sigma :
  \Sub\,\Gamma\,\Delta) \times \Tm_{i,j}\,\Gamma\,(A[\sigma])$, we recover the
  operation which extends a morphism with a term. In the initial model, this
  extends a parallel substitution with an extra term, thus justifying the view
  of substitutions as lists of terms. We denote the extension operation as
  $(\sigma,\,t)$ for $\sigma : \Sub\,\Gamma\,\Delta$ and $t : \Tm_{i,j}\,\Gamma\,(A[\sigma])$.
\end{itemize}

\begin{notation}
De Bruijn indices are rather hard to read, so we will sometimes use nameful notation
for binders and substitutions. For example, we may write $\Gamma \ext (x : A)
\ext (y : B)$ for a context, and subsequently write $B[x \mapsto t]$ for
substituting the $x$ variable for some term $t : \Tm_{i,j}\,\Gamma\,A$. Using
nameless notation, we would instead have $B : \Ty_{i,j}\,(\Gamma \ext A)$ and
$B[\id,\,t] : \Ty_{i,j}\,\Gamma$; here we recover single substitution by extending the identity
substitution $\id : \Sub\,\Gamma\,\Gamma$ with $t$.

We may also leave weakening implicit: if a type or term is in a context
$\Gamma$, we may use it in an extended context $\Gamma \ext A$ without marking
the weakening substitution.
\end{notation}

\begin{definition}
A \textbf{model of 2LTT} is a model of basic 2LTT which supports certain type
formers. For the sake of brevity, we specify here only a small collection of
type formers. However, we will argue later that our results extend to more
general notions of inductive types. We specify type formers in the following. We
omit substitution rules for type and term formers; all of these can be specified in
an evident structural way, e.g.\ as in $(\Lift A)[\sigma] = \Lift(A[\sigma])$.
\begin{itemize}
\item \emph{Universes}. For each $i$ and $j$, we have a Coquand-style universe
  \cite{coquand2018canonicity} in $\Ty_{i,j}$. This consists of $\U_{i,j} : \Ty_{i,j+1}\,\Gamma$,
  together with $\El : \Tm_{i,j+1}\,\Gamma\,\U_{i,j} \to \Ty_{i,j}\,\Gamma$ and $\Code$, where $\Code$
  and $\El$ are inverses.
\item \emph{$\Sigma$-types}. We have $\Sigma\,(x : A)\,B : \Ty_{i,j}\,\Gamma$ for $A : \Ty_{i,j}\,\Gamma$
  and $B : \Ty_{i,j}\,(\Gamma \ext (x : A))$, together with a natural isomorphism consisting of pairing and projections:
  \[ \Tm_{i,j}\,\Gamma\,(\Sigma\,(x : A)\,B) \simeq (t : \Tm_{i,j}\,\Gamma\,A) \times \Tm_{i,j}\,\Gamma\,(B[x\mapsto t]) \]
  We write $(t,\,u)$ for pairing and $\fst$ and $\snd$ for projections.
\item \emph{Function types}.
  We have $\Pi\,(x : A)\,B : \Ty_{i,j}\,\Gamma$ for $A : \Ty_{i,j}\,\Gamma$ and
  $B : \Ty_{i,j}\,(\Gamma \ext (x : A))$, together with abstraction and application
  \begin{alignat*}{3}
    & \lam &&: \Tm_{i,j}\,(\Gamma \ext (x : A))\,B \to \Tm_{i,j}\,\Gamma\,(\Pi\,(x : A)\,B) \\
    & \app &&: \Tm_{i,j}\,\Gamma\,(\Pi\,(x : A)\,B) \to \Tm_{i,j}\,(\Gamma \ext (x : A))\,B
  \end{alignat*}
  such that $\lam$ and $\app$ are inverses. Since we have explicit
  substitutions, this specification of $\app$ is equivalent to the
  ``traditional'' one:
  \[ \app' : (t : \Tm_{i,j}\,\Gamma\,(\Pi\,(x :
  A)\,B))\to (u : \Tm_{i,j}\,\Gamma\,A) \to \Tm_{i,j}\,\Gamma\,(B[x \mapsto
    u]).\]
  The traditional version is definable as $(\app\,t)[\id,\,u]$. We use
  the $\app$-$\lam$ isomorphism because it is formally more convenient.

\item \emph{Natural numbers}. We have $\Nat_{i,j} : \Ty_{i,j}\,\Gamma$, $\zero_{i,j} : \Tm_{i,j}\,\Gamma\,\Nat_{i,j}$,
  and $\suc_{i,j} : \Tm_{i,j}\,\Gamma\,\Nat_{i,j} \to \Tm_{i,j}\,\Gamma\,\Nat_{i,j}$. The eliminator is the following.
  \begin{alignat*}{5}
    & \NatElim :\hspace{0.5em}
                &&  (P &&: \Ty_{i,k}\,(\Gamma \ext (n : \Nat_{i,j}))) \\
    &           &&  (z &&: \Tm_{i,k}\,\Gamma\,(P[n\mapsto \zero_{i,j}])) \\
    &           &&  (s &&: \Tm_{i,k}\,(\Gamma \ext (n : \Nat_{i,j}) \ext (\mit{pn} : P[n\mapsto n]))\,(P[n\mapsto \suc_{i,j}\,n]))\\
    &           &&  (t &&: \Tm_{i,j}\,\Gamma\,\Nat_{i,j}) \\
    &\hspace{3em}\to        &&&& \Tm_{i,k}\,\Gamma\,(P[n\mapsto t]))
  \end{alignat*}
  We also have the $\beta$-rules:
  \begin{alignat*}{5}
    & \NatElim\,P\,z\,s\,\zero_{i,j}     &&= z \\
    & \NatElim\,P\,z\,s\,(\suc_{i,j}\,t) &&= s[n \mapsto t,\,\mit{pn} \mapsto \NatElim\,P\,z\,s\,t]
  \end{alignat*}
  Note that we can eliminate from a level $j$ to any level $k$.
\end{itemize}
\end{definition}

@@

\subsubsection{The setoid interpretation of the syntax}\label{sec:2lttsetoid} We need to show
that the 2LTT syntax, as a particular inductive type, can be modeled in setoids.
Unfortunately, there is no prior work which presents the setoid interpretation
of quotient types in the generality of quotient induction-induction. We sketch
it for our specific use case, but we argue that the interpretation is fairly
mechanical and could be adapted to any quotient inductive-inductive type.
We focus on contexts and types here.
